Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2021-04-05 Manuscript/HI_Model_ms.Rmd
updated manuscript files, added journal cover letter
updated manuscript files

### *4.2 Comparison to other studies*
There are few estimates of contemporary ecosystem carbon balance for the main Hawaiian Islands, and even fewer model projections of future ecosystem carbon balance in response to climate and land use change.  Our mean annual NBP estimate of 0.64 TgC y^-1^ for the period 2011-2020 agrees well with a recent State of Hawaiʻi Greenhouse Gas Inventory report, which estimated an annual net carbon sink of 0.66 Tg C in 2015 from agriculture, forestry, and other land uses [@hawaii_inventory_2019]. In contrast, our NBP estimate for the past decade was ~ 88% higher than a previous statewide LUCAS model estimate covering the same time period [0.341 Tg C y^-1^; @selmants_baseline_2017]. This discrepancy was likely driven by modifications in how we calculated NPP, soil R~h~, and soil aquatic carbon loss compared to previous versions of the LUCAS model, as well our modelʻs finer spatial resolution [@selmants_baseline_2017; @daniel_integrating_2018]. Previous versions of a Hawaiʻi LUCAS model were run at 1-km spatial resolution and simulation cells within each unique combination of moisture zone and state type all had the same mean IBIS-derived NPP value applied to them at the beginning of the simulation period. In contrast, our NPP estimates at 250-m spatial resolution were adjusted on a cell-by-cell basis using Hawaiʻi-specific climate data as described in section 2.3. As a result, our statewide NPP estimates from 2011-2020 were 9.5% lower on average than previous LUCAS model estimates for Hawaiʻi during the same time period [@selmants_baseline_2017], likely because of the greater influence of more arid simulation cells. Soil carbon losses via R~h~, leaching, and overland flow in previous versions of the LUCAS model were calculated as the ratio of the IBIS-derived flux to the size of the originating carbon stock, in this case soil organic carbon to 1-m depth [@daniel_integrating_2018]. Here we calculated soil R~h~ and aquatic carbon losses as the ratio of the mean IBIS-derived flux to the microclimate-adjusted NPP value of each simulation cell, which is a more realistic driver than stock size [@jackson_ecology_2017]. Compared to previous Hawaiʻi LUCAS model estimates [@selmants_baseline_2017], soil R~h~ and aquatic carbon losses from 2011-2020 were reduced by an average of 15% and 21%, respectively, which widened the gap between between carbon gain (NPP) and carbon losses and accounted for the overall increase in NBP estimates for this time period.

@@

Our model does not currently differentiate between forests dominated by native versus non-native tree species, which could influence estimates of ecosystem carbon balance. Native forests in the Hawaiian Islands are dominated by ʻōhiʻa (_Metrosideros polymorpha_ Gauditch), an endemic foundational tree species found across a broad range of climatic and edaphic conditions [@ziegler_hawaiian_2002]. Beginning in 2010, a fungal disease termed Rapid ʻŌhiʻa Death (ROD) caused by two _Ceratocystus_ spp. has emerged and caused widespread mortality to mature ʻōhiʻa trees across a range of size classes, primarily on Hawaiʻi Island [@mortenson_assessing_2016; @fortini_evolving_2019]. The distribution and potential range of this emerging threat to Hawaiʻi's dominant native tree species have only recently been mapped [@vaughn_approach_2018; @fortini_evolving_2019], but the pulse of newly dead organic matter and reduction in photosynthetic capacity induced by widespread tree mortality could significantly alter ecosystem carbon balance over the long-term [@sleeter_effects_2019]. ROD-affected forests could also undergo a replacement of canopy dominant stress-tolerating ʻōhiʻa trees by non-native tree species with faster relative growth rates, lower wood density, and faster tissue turnover, potentially alter the long-term trajectory of forest carbon cycling in Hawaiʻi. New model projections for Hawaiʻi that incorporate ROD spread rates and forest restoration scenarios will therefore require differentiation among forest ecosystems dominated by native and non-native tree species.

@@

This study was funded by the U.S. Geological Survey Biological Carbon Sequestration Program and the Pacific Islands Climate Adaptation Science Center. Thanks to Leonardo Frid and Colin Daniel of ApexRMS for assistance with SyncroSim software, and to Nicholas Koch of Forest Solutions, Inc. for information on _Eucalyptus_ spp. harvesting in Hawaiʻi. Thanks also to Christian Giardina and Zhiliang Zhu for providing the initial impetus for this research. Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.

@@

>>>>

### *4.2 Comparison to other studies*
There are few estimates of contemporary ecosystem carbon balance for the main Hawaiian Islands, and even fewer model projections of future ecosystem carbon balance in response to climate and land use change.  Our mean annual NBP estimate of 0.64 TgC y^-1^ for the period 2011-2020 agrees well with a recent State of Hawaiʻi Greenhouse Gas Inventory report, which estimated an annual net carbon sink of 0.66 Tg C in 2015 from agriculture, forestry, and other land uses [@hawaii_inventory_2019]. In contrast, our NBP estimate for the past decade was ~ 88% higher than a previous statewide LUCAS model estimate covering the same time period [0.341 Tg C y^-1^; @selmants_baseline_2017]. This discrepancy was likely driven by modifications in how we calculated NPP, soil R~h~, and soil aquatic carbon loss compared to previous versions of the LUCAS model, as well our model's finer spatial resolution [@selmants_baseline_2017; @daniel_integrating_2018]. Previous versions of a Hawaiʻi LUCAS model were run at 1-km spatial resolution and simulation cells within each unique combination of moisture zone and state type all had the same mean IBIS-derived NPP value applied to them at the beginning of the simulation period. In contrast, our NPP estimates at 250-m spatial resolution were adjusted on a cell-by-cell basis using Hawaiʻi-specific climate data as described in section 2.3. As a result, our statewide NPP estimates from 2011-2020 were 9.5% lower on average than previous LUCAS model estimates for Hawaiʻi during the same time period [@selmants_baseline_2017], likely because of the greater influence of more arid simulation cells. Soil carbon losses via R~h~, leaching, and overland flow in previous versions of the LUCAS model were calculated as the ratio of the IBIS-derived flux to the size of the originating carbon stock, in this case soil organic carbon to 1-m depth [@daniel_integrating_2018]. Here we calculated soil R~h~ and aquatic carbon losses as the ratio of the mean IBIS-derived flux to the microclimate-adjusted NPP value of each simulation cell, which is a more realistic driver than stock size [@jackson_ecology_2017]. Compared to previous Hawaiʻi LUCAS model estimates [@selmants_baseline_2017], soil R~h~ and aquatic carbon losses from 2011-2020 were reduced by an average of 15% and 21%, respectively, which widened the gap between between carbon gain (NPP) and carbon losses and accounted for the overall increase in NBP estimates for this time period.

@@

Our model does not currently differentiate between forests dominated by native versus non-native tree species, which could influence estimates of ecosystem carbon balance. Native forests in the Hawaiian Islands are dominated by ʻōhiʻa (_Metrosideros polymorpha_ Gauditch), an endemic foundational tree species found across a broad range of climatic and edaphic conditions [@ziegler_hawaiian_2002]. Beginning in 2010, a fungal disease termed Rapid ʻŌhiʻa Death (ROD) caused by two _Ceratocystus_ spp. has emerged and caused widespread mortality to mature ʻōhiʻa trees across a range of size classes, primarily on Hawaiʻi Island [@mortenson_assessing_2016; @fortini_evolving_2019]. The distribution and potential range of this emerging threat to Hawaiʻi's dominant native tree species have only recently been mapped [@vaughn_approach_2018; @fortini_evolving_2019], but the pulse of newly dead organic matter and reduction in photosynthetic capacity induced by widespread tree mortality could significantly alter ecosystem carbon balance over the long-term [@sleeter_effects_2019]. ROD-affected forests could also undergo a replacement of canopy dominant stress-tolerating ʻōhiʻa trees by non-native tree species with faster relative growth rates, lower wood density, and faster tissue turnover, potentially altering the long-term trajectory of carbon cycling in Hawaiian forest ecosystems. New model projections for Hawaiʻi that incorporate ROD spread rates and forest restoration scenarios will therefore require differentiation among forest ecosystems dominated by native and non-native tree species.

@@

This study was funded by the U.S. Geological Survey Biological Carbon Sequestration Program and the Pacific Islands Climate Adaptation Science Center. Thanks to Leonardo Frid and Colin Daniel of ApexRMS for assistance with SyncroSim software, and to Nicholas Koch of Forest Solutions, Inc. for information on _Eucalyptus_ spp. harvesting in Hawaiʻi. Thanks also to Christian Giardina and Zhiliang Zhu for providing the impetus for this research. Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.

@@

Clay Trauernicht https://orcid.org/0000-0002-1509-8536
